The Influence of Connectivity on Dung Beetle Communities

### 2025-07-06 corridor_docs/eric_ms_thesis/MS_thesis.Rmd
the end pt 1

As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented [@diaz_pervasive_2019; @ma_global_2023; @taubert_global_2018]. Like many ecological processes, fragmentation is a complex and multifaceted phenomenon bringing about many consequences which can be both positive and negative for ecosystems [@fahrig_effects_2003; @fletcher_is_2018]. However, as habitats are broken down community structures are significantly altered [@WOS:000419965700012; @haddad_corridor_2003; @jennings_composition_2006]. This alteration of structure typically lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015].

@@

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions [@hasan_dung_2024]. The removal, breakdown, and burial of animal feces drive important ecosystem interactions provided by dung beetles enhancing nutrient cycling and soil quality, the reduction of breeding sites for parasites, and a reduction in methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013; @nichols_ecological_2008]. [**EB edit:** I added the following to justify dung veetles more, and then made the questions a second paragraph] Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits [e.g., size, foraging style, resource-use, [@decastro-arrazola_trait-based_2023; @ospina-garces_dung_2018]]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

When comparing the total abundance of beetles (all species combined) across patch types, there were no significant difference with one exception: fewer individuals were captured in rectangular patches($\beta$ = -0.402, P \< 0.001, Table 4).

Statistical analysis of abundance focused on the six most abundant species. A generalized linear mixed model identified significant effects for both species ID and patch type on dung beetle abundance (Table 7). The baseline abundance corresponds to the abundance of *Aphodius alloblackburneus* in connected patches. Compared to this baseline, *Canthon vigilans*, *Ateuchus lecontei*, and *Phanaeus igneus* showed significantly higher abundance ($\beta$ = 1.05, 0.73, 0.66 respectively; all p \< 0.001), while *Onthophagus pennsylvanicus* had lower abundance ($\beta$ = –0.80, p \< 0.001).

Species-patch interactions indicated species specific responses to treatments. In matrix plots, the positive effect on abundance was reduced for all other species (e.g., cvig × Matrix $\beta$ = –0.59, p \< 0.001). *Ateuchus lecontei* and *Phanaeus igneus* showed increased effect on abundance in rectangle patches relative to expectations ($\beta$ = 0.40 and 0.47, respectively; p \< 0.05). Random effects indicated moderate variability among blocks (standard deviation of intercept = 0.51).

@@

Plotting species richness by patch type reveals consistent richness across patch types with some variation between sampling blocks (Figure 5).The number of species per patch varied from N = 8 (rectangle patch in block 8) to N = 13 (matrix patch in block 53N). Modeling the effect of patch type on species richness with block as a random effect determined there was no significant differences among patch types. Comparing treatments using connected patches as a baseline resulted in no significant differences in matrix ($\beta$ = 0.05, p = 0.83), rectangle ($\beta$ = -0.07, p = 0.74), and winged patches ($\beta$ = -0.05, p = 0.83).

For Shannon Diversity, the estimated mean in Connected patches was 5.29 (SE = 0.66, t = 7.95). None of the alternative patch types showed statistically significant differences compared to Connected: Matrix ($\beta$ = –0.085, SE = 0.50, t = –0.17), Rectangle ($\beta$ = 0.15, SE = 0.50, t = 0.30), or Winged ($\beta$ = –0.34, SE = 0.50, t = –0.68). Similarly, for Simpson's Diversity, the average value in Connected patches was 4.07 (SE = 0.64, t = 6.37). Again, none of the other patch types had significant effects: Matrix ($\beta$ = 0.11, SE = 0.41, t = 0.26), Rectangle ($\beta$ = 0.17, SE = 0.41, t = 0.40), and Winged ($\beta$ = –0.37, SE = 0.41, t = –0.91).

@@

This study advances our understanding of the factors shaping dung beetle community composition in temperate regions of the southeastern United States. In addition, the experimental design enables direct comparisons between populations in continuous matrix habitat and those in both isolated and corridor connected patches. Our main findings emphasized: (1) Habitat type and patch shape were the main driving factors for dung beetle species abundances were composed, however effects were species specific. (2) Patch shape and isolation had less of an influence on species richness which was relatively even on both a patch and block level. (3) Species diversity metrics were also relatively even across patch types however varied widely by sampling blocks. These results suggest that while there may indeed be effects of patch structure and connectivity on dung beetle abundances and community composition, other landscape scale drivers appear to be more prominent for species richness and diversity.

@@

In comparing patc

@@

>>>>

As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented [@diaz_pervasive_2019; @ma_global_2023; @taubert_global_2018]. Like many ecological processes, fragmentation is a complex and multifaceted phenomenon bringing about many consequences which can be both positive and negative for ecosystems [@fahrig_effects_2003; @fletcher_is_2018]. However, as habitats are broken down community structures are significantly altered [@WOS:000419965700012; @haddad_corridor_2003; @jennings_composition_2006; **harrison and bruna**]. This alteration of structure typically lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015].

@@

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions [@hasan_dung_2024]. The removal, breakdown, and burial of animal feces drive important ecosystem interactions provided by dung beetles enhancing nutrient cycling and soil quality, the reduction of breeding sites for parasites, and a reduction in methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013; @nichols_ecological_2008]. [**EB edit:** I added the following to justify dung beetles more, and then made the questions a second paragraph] Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits [e.g., size, foraging style, resource-use, [@decastro-arrazola_trait-based_2023; @ospina-garces_dung_2018]]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

When comparing the overall abundance of beetles (all species combined) across patch types, matrix plots had the highest captures, followed by connected patches, then winged, with the fewest in rectangular patches. Abundances from connected patches were not significantly different from those in the matrix while rectangle and winged patches had significantly fewer than the matrix ($\beta$ = -0.528 and -0.303 respectively, P \< 0.001, Table 4).

Statistical analysis of abundance focused on the six most abundant species. A generalized linear mixed model identified significant effects for both species ID and patch type on dung beetle abundance (Table 7). The baseline abundance corresponds to the abundance of *Aphodius alloblackburneus* in matrix patches. Compared to this baseline, results were highly variable, emphasizing species specific responses to patch type. Similarly, species showed to have disproportionate responses to patch since interaction terms varied widely (Table 6).

@@

Plotting species richness by patch type reveals consistent richness across patch types with some variation between sampling blocks (Figure 5).The number of species per patch varied from N = 8 (rectangle patch in block 8) to N = 13 (matrix patch in block 53N). Modeling the effect of patch type on species richness with block as a random effect determined there was no significant differences among patch types. Comparing treatments using matrix patches as a baseline resulted in no significant differences in connected ($\beta$ = 0.216, p = 0.839; Table 7), rectangle ($\beta$ = -0.548, p = 0.583; Table 7), and winged patches ($\beta$ = -0.095, p = 0.663; Table 7).

Biodiversity was also even among patch types, however metric values were varied by sampling blocks. For Shannon Diversity, the estimated mean in matrix patches was 5.202 (SE = 0.665, t = 7.825). None of the alternative patch types showed statistically significant differences compared to matrix: connected ($\beta$ = –0.085, SE = 0.495, t = 0.17), rectangle ($\beta$ = 0.234, SE = 0.495, t = 0.472), or winged ($\beta$ = –0.251, SE = 0.495, t = –0.506). Similarly, for Simpson's Diversity, the average value in matrix patches was 4.181 (SE = 0.640, t = 6.535). Again, none of the other patch types had significant effects: matrix ($\beta$ = -0.107, SE = 0.411, t = -0.259), rectangle ($\beta$ = 0.059, SE = 0.411, t = 0.143), and winged ($\beta$ = –0.480, SE = 0.411, t = –1.169).

@@

This study advances our understanding of the factors shaping dung beetle community composition in temperate regions of the southeastern United States. In addition, the experimental design enables direct comparisons between populations in continuous matrix habitat and those in both isolated and corridor connected patches. Our main findings emphasized: (1) Habitat type and patch shape were the main driving factors for determining how dung beetle species abundances were composed, however effects were species specific. (2) Patch shape and isolation had less of an influence on species richness which was relatively even on both a patch and block level. (3) Species diversity metrics were also relatively even across patch types however varied widely by sampling blocks. These results suggest that while there may indeed be effects of patch structure and connectivity on dung beetle abundances and community composition, other landscape scale drivers appear to be more prominent for species richness and diversity.

@@

In our comparison of dung beetle compositions, connectivity and habitat edge accounted for differences in dung beetle abundances, yet species richness and diversity were even across patch types and in  the matrix. Total beetle counts were consistently highest in the matrix, in comparison abundances is connected patches were not significantly different while those of the winged and rectangular patches were lower. Additionally, we also observed that patch effects were not equally proportional for all species. This suggests that the species captured could be exhibiting habitat preferences between the open patches versus forested matrix. Another potential explanation is that species in the matrix are acting as a source population which feeds into patches with edge acting as a drift fence for directing movement. The latter example seems more likely because flies, which have similar resource dependencies were also found to have a similar interaction ** fried **, and also studies on dung beetles in the tropics found more dramatic differences in dung beetle populations in fragments and matrix [@bustamante-sanchez_dung_2004; @WOS:000180797500010; @WOS:000288810500026].

@@

Although richness and diversity were the same among treatments, there was notable variability between sampling blocks.
